<div id="table-of-contents">
<h2>Table of Contents</h2>
<div id="text-table-of-contents">
<ul>
<li><a href="#orgheadline1">1. Introduction</a></li>
<li><a href="#orgheadline7">2. Usage</a>
<ul>
<li><a href="#orgheadline2">2.1. Get miRNA-target gene interactions for a specific gene</a></li>
<li><a href="#orgheadline3">2.2. Get miRNA-target gene interactions for a miRNA</a></li>
<li><a href="#orgheadline4">2.3. Get grouped miRNA-target gene interactions</a></li>
<li><a href="#orgheadline5">2.4. Alternative way to fetch data from the database</a></li>
<li><a href="#orgheadline6">2.5. Conversions between miRNA identifiers</a></li>
</ul>
</li>
<li><a href="#orgheadline8">3. Database layout</a></li>
<li><a href="#orgheadline9">4. References</a></li>
</ul>
</div>
</div>

---
title: "mirtarbase: a database of validated miRNA target gene interactions"
graphics: yes
output:
  BiocStyle::html_document:
    toc_depth: 2
  BiocStyle::pdf_document:
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{mirtarbase: a database of validated miRNA target gene interactions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{mirtarbase}
  %\VignettePackage{mirtarbase}
  %\VignetteKeywords{annotation,database,miRNA}
bibliography: references.bib
csl: biomed-central.csl
references:
- id: dummy
  title: no title
  author:
  - family: noname
    given: noname
---

**Authors**: `r packageDescription("mirtarbase")$Author`<br />
**Modified**: 15 February, 2016<br />
**Compiled**: `r date()`

```{r echo=FALSE, results='asis'}
BiocStyle::markdown()
```

# Introduction<a id="orgheadline1"></a>

The `mirtarbase` package provides the experimentally validated miRNA-target gene
interactions (MTIs) defined in the miRTarbase database [@Hsu:2014co].

The database, which was build based on the Excel spread sheet that can be
downloaded from the miRTarbase main site <http://mirtarbase.mbc.nctu.edu.tw/>, is
automatically loaded upon library startup and bound to the environment variable
`MirtarbaseDb.v<release>`, where `<release>` stands for the release of the
miRTarBase. In addition, a shortcut `mirtarbase` is automatically generated and
linked to the (most recent) release available in the package.

```{r warning=FALSE}
library(mirtarbase)
```

# Usage<a id="orgheadline7"></a>

After the package is loaded the database can be accessed through the
`MirtarbaseDb` object bound to the `mirtarbase` variable.

```{r }
## print some information for the package
mirtarbase
```

The miRTarbase defines a MTI for each mature miRNA - target gene pair and
provides at least one publication in which this interaction was experimentally
verified along with the support type (i.e. one of four evidence grades that was
defined by the developers), the experiments supporting the interaction and the
Pubmed ID. This information is stored internally in a SQLite database (see
Section [3](#orgtarget1) for the layout of the database and the
attributes/columns).

Before we start with some usage scenarios for the package it is important to
understand the mechanism to fetch specific values from the database, i.e. how
the results can be filtered. The package uses the same filtering system than the
`ensembldb` Bioconductor package and extends it by some additional filters.
The following filter classes are used by the package (in alphabetical order):

-   `EntrezidFilter`: filter the results based on the target genes' NCBI
    Entrezgene identifiers.
-   `ExperimentFilter`: restrict the results to specific experiments (use
    `listExperiments(mirtarbase)` to get a list of all experiments).
-   `GenenameFilter`: filter the results based on the target genes' names
    (symbols).
-   `MatmirnaFilter`: search for MTIs for specific mature miRNA names
    (e.g. *hsa-miR-15b-5p*).
-   `MarmirnaidFilter`: search for MTIs for specific mature miRNAs identified by
    their mirbase accession number.
-   `MirfamFilter`: filter the results based on miRNAs of specific miRNA families.
-   `MirfamidFilter`: filter the results based on miRNAs of specific miRNA
    families identified by their mirbase accession number.
-   `PremirnaFilter`: filter the results for MTIs of mature miRNAs derived from
    the specified pre-miRNA(s).
-   `PremirnaidFilter`: filter the results for MTIs of mature miRNAs derived from
    the specified pre-miRNA(s) identified by their mirbase accession number.
-   `PublicationFilter`: filter the results based on the publication (identified
    by their Pubmed IDs) in which the MTI was described.
-   `SpeciesFilter`: filter the results for target genes or miRNAs of the
    specified species (use `listSpecies(mirtarbase, "gene")` or
    `listSpecies(mirtarbase, "mirna")`).
-   `SupportTypeFilter`: filter the results based on the evidence/support type of
    the MTI (use `listSupportTypes(mirtarbase)` to get a list of allowed values
    for support type).

These filters can be used individually or can be combined to generate more
specific queries. Furthermore, the parameter `condition` allows some more
flexibility to choose which entries should be fetched. `condition` can take the
following values: `=`, `!=`, `like`, `not like` with the latter two allowing
partial matching. In addition, each filter can take a single or multiple values
(see examples in the next section).

## Get miRNA-target gene interactions for a specific gene<a id="orgheadline2"></a>

As an example we want to retrieve all MTIs for the gene *BCL2*, i.e. we want to
get all (mature) miRNAs that have been shown to target this gene. To this end we
define a `GenenameFilter` with the value `BCL2` and submit this to the `mtis`
call. As a result we get a `MTIList`, which is essentially a `list` of `MTI`
objects that describe the interaction. Each interaction is defined by the mature
miRNA name and the name of the target gene (accessible through the `matmirna`
and `gene` methods, respectively) as well as the collection of evidences for the
interaction. These evidences, or rather publications, are accessible through the
`reports` method and specify by which experimental methods the interaction was
validated and in which publication this interaction has been described. The
curators of the miRTarbase manually assigned each evidence one of four *support
types* which is accessible through the `supportedBy` method.

The code below simply fetches all MTIs for the gene *BCL2* from the database.

```{r }
## Query the database to fetch all MTIs for the target gene BCL2
BCL2 <- mtis(mirtarbase, filter=list(GenenameFilter("BCL2")))
BCL2

## To print some more information on a single MTI
BCL2[[1]]

## How many interactions did we get?
length(BCL2)

## These are however of all species as we did not specify a species filter
## and miRTarBase lists interactions for all species.
sort(table(mirnaSpecies(BCL2)), decreasing=TRUE)
```

In order to restrict the MTIs to human genes and human miRNAs it is advisable to
add one or more `SpeciesFilter` to the query.

```{r }
## We can use the listSpecies method to get the names of all supported species
## from the database:
sort(listSpecies(mirtarbase))

## We want to get all human mature miRNAs that target human gene BCL2
BCL2 <- mtis(mirtarbase, filter=list(GenenameFilter("BCL2"),
				     SpeciesFilter("Homo sapiens", feature="gene"),
				     SpeciesFilter("Homo sapiens", feature="mirna")))

## Now we have only human miRNAs. We can now make a table of the miRNA,
## the support type and the number of publications for each MTI
BCL2.df <- data.frame(miRNA=matmirna(BCL2),
		      reports=reportCount(BCL2),
		      support_type=unlist(lapply(supportedBy(BCL2), function(z){
			  return(paste(unique(z), collapse=";"))
		      })))

## Display the MTIs described by the most publications
head(BCL2.df[order(BCL2.df$reports, decreasing=TRUE), ])
```

So, there is evidence that e.g. miR-16-5p is targeting the gene BCL2, along with
miR-15a-5p. We can also enrich this table with the information of the
pre-miRNA(s) in which the mature miRNA is encoded. In addition, we can group the
miRNAs also by the miRNA family. Note that each mature miRNA can be eventually
encoded in more than one pre-miRNA, each mature miRNA (and each pre-miRNA) is
supposed to be part of one miRNA family.

```{r }
BCL2.df <- cbind(BCL2.df,
		 premirna=unlist(lapply(BCL2, function(z){
		     return(paste(premirna(z), collapse=";"))
		 })),
		 mirfam=mirfam(BCL2))

## Note: there are some mature miRNAs that can not be mapped to pre-miRNA
## or mirfam names.
sum(is.na(as.character(BCL2.df$mirfam)))

## the miRNA with most evidences (miR-16-5p) is actually encoded in two
## precursors:
premirna(BCL2$MIRT001800)

## The miRNA families from which most miRNAs target BCL2 are listed below:
sort(table(as.character(BCL2.df$mirfam)), decreasing=TRUE)

## The miRNAs from the mir-15 family targeting BCL2 are
MTI.mir15 <- BCL2[ which(unlist(lapply(BCL2, mirfam))=="mir-15") ]
## the mature miRNAs from this family:
MTI.mir15

## Extract the mature miRNA IDs
matmirna(MTI.mir15)
## And the pre-miRNAs:
premirna(MTI.mir15)
```

The missing mapping of mature miRNAs to pre-miRNA names or mirfam identifiers
observed above is in many instances caused by different mirbase versions on
which the `mirbase.db` package and the miRTarbase bases. In addition, not all
mature miRNAs are annotated to miRNA families.

As we have seen above, we can use the methods `matmirna`, `premirna` and
`mirfam` on `MTI` or `MTIList` objects to retrieve the mature miRNA involved in
the miRNA-target gene interaction, the pre-miRNA in which the mature miRNA is
encoded and the miRNA family to which the pre-miRNA(s) belong.

## Get miRNA-target gene interactions for a miRNA<a id="orgheadline3"></a>

Next we retrieve MTIs between miRNAs of the mir-15 family and genes which names
start with *BCL2*. For this we define a `GenenameFilter` with `"like"` as
condition and a pattern for the gene name.

```{r }
## Get all miRNA-target gene interactions betwee mature miRNAs from the
## mir-15 family and genes starting with BCL2
BCLs <- mtis(mirtarbase, filter=list(MirfamFilter("mir-15"),
				     GenenameFilter("BCL2%", condition="like"),
				     SpeciesFilter("Homo sapiens"))
	    )
BCLs
```

According to this information the miRNA *miR-195-5p* targets both, a pro- and an
anti-apoptotic member of the *BCL2* gene family (*BCL2L11* and *BCL2*,
respectively).

By default, the results are returned by the `mtis` method as `MTIList` object,
but we could also specify `"data.frame"` as the `return.type` to retrieve the
data as `data.frame`. This allows to retrieve only specific information from the
database by specifying the columns that should be returned.

```{r }
onlyGeneNames <- mtis(mirtarbase, filter=list(MirfamFilter("mir-15"),
					      GenenameFilter("BCL2%", condition="like"),
					      SpeciesFilter("Homo sapiens")),
		      columns=c("mirna", "target_gene"), return.type="data.frame")
head(onlyGeneNames)
```

Also members of the *mir-17* family have been reported to target genes from the
*BCL2* gene family [@Ventura:2008gk], thus we retrieve next all MTIs between miRNAs of the miRNA
families *mir-15* or *mir-17* and some of the genes from the BCL2 gene family, a
gene family involved in, and regulating, the intrinsic apoptotic pathway.

To retrieve values for more than one gene, respectively miRNA family, we can
submit a character vector of the respective ids to the filters.

```{r }
## retrieving all MTIs between miRNAs from the mir-15 and mir-17 families
## and some genes from the BCL2 gene family
BCLs <- mtis(mirtarbase,
	     filter=list(MirfamFilter(c("mir-15", "mir-17")),
		 GenenameFilter(c("BCL2", "BCL2L11", "PMAIP1", "MCL1")),
		 SpeciesFilter("Homo sapiens"))
	    )
BCLs
## the miRNA - gene pairs:
data.frame(miRNA=matmirna(BCLs),
	   gene=gene(BCLs),
	   report_count=reportCount(BCLs))
```

Apparently, miRNAs from both the miR-15 and the miR-17 family target genes of
the BCL2 gene family and are thus also involved in the regulation of the
apoptotic pathway.

Next we evaluate the evidence grades of the interaction and remove all MTIs that
are not of the *Functional MTI* support type (the type with the highest evidence
grade).

```{r }
funcMti <- unlist(lapply(BCLs, function(z){
    return(any(supportedBy(z)=="Functional MTI"))
}))
sum(funcMti)
length(funcMti)

## We could now use this logical vector to sub-set the list.
## Alternatively, we can also re-perform the query and fetch only interactions of that
## support type, which has the advantage that also only the publications of the
## corresponding support type are loaded.
BCLs <- mtis(mirtarbase,
	       filter=list(MirfamFilter(c("mir-15", "mir-17")),
		   GenenameFilter(c("BCL2", "BCL2L11", "PMAIP1", "MCL1")),
		   SpeciesFilter("Homo sapiens"),
		   SupportTypeFilter("Functional MTI"))
	    )
## the miRNA - gene pairs:
data.frame(miRNA=matmirna(BCLs),
	   gene=gene(BCLs),
	   report_count=reportCount(BCLs)
	  )
```

This considerably reduced the list of interactions and also decreased the number
of reports per MTI.

## Get grouped miRNA-target gene interactions<a id="orgheadline4"></a>

Sometimes it might be useful to group the miRNA-target gene interactions by some
factor, e.g. by genes or miRNAs. The method `mtisBy` allows to fetch `MTIs`
grouped by any column from the database. It is possible to group the results
by gene, (mature miRNA), entrezid, support type, Pubmed ID, pre-miRNA name,
miRFam name or by species. The result will be a `list` with the names being the
factor by which the interactions are grouped and each element being a `MTIList`
of the MTIs.

In the example below we fetch all MTIs for the genes *BCL2*, *BCL2L11*, *MCL1*
and group them by miRNA family.

```{r }
Filters <- list(SpeciesFilter(c("Homo sapiens")),
		GenenameFilter(c("BCL2", "BCL2L11", "MCL1")))

BCL2by <- mtisBy(mirtarbase, filter=Filters, by="mirfam")
head(BCL2by)
```

In a similar way we can also fetch the data grouped by gene.

```{r }
BCL2by <- mtisBy(mirtarbase, filter=Filters, by="gene")
BCL2by
```

## Alternative way to fetch data from the database<a id="orgheadline5"></a>

By default, the `mtis` method returns a list of `MTI` objects (`MTIList`) which
is sufficient for most use cases. Alternatively, however, the `mtis` method can
also return the results as a `data.frame`. In addition to a significant
performance improvement this also enables to select only specific columns
from the database. Note however that by default the method returns all
columns from the database which results in a `data.frame` with one
MTI-publication per row, i.e. the same MTI represented by the miRNA-gene pair
can be present in many rows of this `data.frame` depending in how many
publications this interaction was identified.

```{r }
## We perform the same call as above, but restrict the information to some selected
## columns and specify to return the results as a data.frame rather than a list
## of MTI objects.
BCLs.df <- mtis(mirtarbase,
		filter=list(MirfamFilter(c("mir-15", "mir-17")),
		    GenenameFilter(c("BCL2", "BCL2L11", "PMAIP1", "MCL1")),
		    SpeciesFilter("Homo sapiens"),
		    SupportTypeFilter("Functional MTI")),
		columns=c("mirna", "target_gene"),
		return.type="data.frame")

BCLs.df
```

## Conversions between miRNA identifiers<a id="orgheadline6"></a>

The `mirtarbase` package provides also methods and functions that allow to map
mature miRNAs to their precursors or to miRNA families. These functions are
essentially wrapper functions that use the information of the `mirbase.db`
Bioconductor package for the conversion. However, since the `mirtarbase` and
`mirbase.db` functions might provide information from different releases, some
of the mappings might not be available. For a complete list of conversion
function refer to the help page of the e.g. `premirna2matmirna` function.

```{r }
## map from pre-miRNA name to mature miRNA name. The function returns by default
## a data.frame
premirna2matmirna(c("hsa-mir-16-1", "hsa-mir-16-2"))

## the same information but as a list:
premirna2matmirna(c("hsa-mir-16-1", "hsa-mir-16-2"), return.type="list")
```

# Database layout<a id="orgtarget1"></a><a id="orgheadline8"></a>

The database consists of 3 tables, `mirtarbase` which contains all information
stored in the xls file from the miRTarBase web site, `pubmed_corpus`, that
contains the content of the *MTI-PubMed\_corpus.txt* file from the miRTarBase web
site and `metadata` with some internal informations. The column names and their
properties are listed below. Each line in the table represents the MTI for a
miRNA and one of its target genes as reported in a publication. Thus, an
interaction between a miRNA and its target gene can be listed in more than one
row, depending on the number of publications it was validated.

-   **mirtarbase**:
    -   `mirtarbase_id`: identifier for the miRNA target gene interaction (MTI). Note
        that this ID is not unique, i.e. MTIs reported in several publications have
        the same ID but are listed in several rows of the table.
    -   `mirna`: mature miRNA name (a.k.a miRNA ID, e.g. *hsa-miR-20a-5p*).
    -   `species_mirna`: the species of the miRNA (e.g. *Homo sapiens*).
    -   `target_gene`: the official gene name (symbol) for the gene (e.g. *DUSP6*, or
        *ush*).
    -   `target_gene_entrez_gene_id`: the NCBI Entrezgene ID for the target gene;
        either `NA` or the (numerical) Entrezgene ID. Contains only unique values, no
        multiple IDs collapsed by any separator.
    -   `species_target_gene`: the species of the target gene.
    -   `experiments`: the experiments providing the evidence for the interaction as
        reported in one publication.
    -   `support_type`: the different types of support (from weak to strong).
    -   `references_pmid`: the Pubmed ID of the publication reporting the MTI. Each
        line with a single Pubmed ID, no empty (`NA`) values.

-   **pubmed\_corpus**
    -   `pmid`: the PubMed ID of the paper describing the MTI.
    -   `key`: the key of the entry (can be *Abstract*, *Title*,
        *Experiment\_method*, *miRNA* or *Target\_gene*).
    -   `value`: the value for the key.

-   **metadata**
    -   `name`: the names of the keys.
    -   `value`: the value for the key.

# References<a id="orgheadline9"></a>